---
title: "Reproducible Research: Peer Assessment 1"
author: Philip Mateescu
output: 
  html_document:
    keep_md: true
---
```{r echo=F}
fd <- function(n, d=2) {
    ## format general number
    format(n, big.mark = ',', big.interval = 3, digits = d)
}
fn <- function(n) {
    ## format general number
    fd(n, d=0)
}
```

### Preamble; Codebook

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

The data for this assignment can be downloaded from the course web
site:

* Dataset: [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) [52K]

The variables included in this dataset are:

* **steps**: Number of steps taking in a 5-minute interval (missing
    values are coded as `NA`)
* **date**: The date on which the measurement was taken in YYYY-MM-DD
    format
* **interval**: Identifier for the 5-minute interval in which
    measurement was taken

We will be using the *dplyr* library for data processing, and `ggplot2` for drawing:
```{r message = F}
library(dplyr)
library(ggplot2)
```


## Loading and preprocessing the data
Fortunately, this repository already contains the **activity.zip** file mentioned above.
However, let's guard against the file being missing for whatever reason:

```{r}
if(!file.exists("activity.zip")){
  download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", destfile="activity.zip")
}
```

We don't need to extract the zip - the `unz` file can do this for us:

```{r}
step.data <- read.csv(unz('activity.zip', filename = 'activity.csv'), colClasses = c('integer', 'Date', 'integer'))
```

For convenience, we'll also load it into a *dplyr* data frame tbl:
```{r}
step.data <- tbl_df(step.data)
step.data
```

## What is mean total number of steps taken per day?

```{r}
mean.steps.per.day <- mean(step.data$steps, na.rm = TRUE)
median.steps.per.day <- median(step.data$steps, na.rm = TRUE)

days.with.steps <- step.data  %>% 
    filter(!is.na(steps)) %>% 
    arrange(date) %>%
    mutate(month.name=months(date), weekday=weekdays(date))
print(days.with.steps)    

totalsteps.per.day <- days.with.steps  %>% 
    group_by(date) %>%
    summarize(steps.total=sum(steps))
# print (totalsteps.per.day)
sum.steps.total <- sum(totalsteps.per.day$steps.total)
mean.steps.total <- mean(totalsteps.per.day$steps.total)
median.steps.total <- median(totalsteps.per.day$steps.total)
```
**Q1: Make a histogram of the total number of steps taken each day**

``` {r}
ggplot(days.with.steps, aes(x = date, y = steps, fill = weekday)) + 
    geom_histogram(stat = 'identity') +
    ## add lines for the mean and the median - notice that being so close together they overlap
    geom_hline(aes(yintercept=mean.steps.total), color = 'red', linetype='dashed') +
    geom_hline(aes(yintercept=median.steps.total), color = 'black', linetype='dotted') +
    labs(title = 'Total number of steps taken each day', x = 'Date', y = 'Steps')
```

**Q2: Calculate and report the mean and median total number of steps taken per day**

Took a **total** of `r fn(sum.steps.total)` steps,
with an **average** of `r fd(mean.steps.total)` steps 
and a **median** value of `r fd(median.steps.total)` steps.


## What is the average daily activity pattern?


**Q1: Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)**

```{r}
## average the time by interval
average.by.time <- days.with.steps %>%
    group_by(interval) %>%
    summarize(steps.average=mean(steps))

# print(head(average.by.time, n = 5))
# print(tail(average.by.time, n = 5))

# This is what we're asked to do, but the default plotting looks... plain.
# plot(average.by.time$interval, average.by.time$steps.average, type='l')
ggplot(average.by.time, aes(x=interval, y=steps.average, color = interval)) +
    geom_line() +  # line plot
    theme(legend.position='none') +  # remove legend - I want the gradient but not the legend
    labs(title='Average number of steps each 5-minute interval across all days',
         x='5-minute intervals',
         y='Average steps taken')
```


**Q2: Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?**

``` {r}
# sort the average.by.time in descending order by steps.average and take the first result
# it'll be the max
average.by.time.sorted <- average.by.time %>%
    arrange(desc(steps.average)) 
```

The **`r average.by.time.sorted[1,]$interval` interval** has the **largest number of steps**:
`r average.by.time.sorted[1,]$steps.average`.

## Imputing missing values


**Q1: Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)**

```{r}
nrow.step.data <- nrow(step.data)
nrow.days.with.steps <- nrow(days.with.steps)
```

Out of *total* `r nrow.step.data` rows, `r nrow.days.with.steps` records *have values*, leaving us with
`r nrow.step.data - nrow.days.with.steps` records **without values**.

**Q2: Devise a strategy for filling in all of the missing values in the dataset.**

Average by time seems a more reliable replacement. Since days might be missing altogether from the dataset, 
any date-relative strategies could yield missing values by themselves.

**Strategy**: we will replace all the missing values with the average for the interval across all days.

**Q3: Create a new dataset that is equal to the original dataset but with the missing data filled in.**

```{r}
# create a copy
step.data2 <- step.data
for(i in 1:nrow.step.data) {
    if (is.na(step.data2$steps[i])) {
        step.data2$steps[i] <- average.by.time[which(average.by.time$interval == step.data$interval[i]),]$steps.average
    }
}
# prove that the copy is different than the original
head(step.data)
head(step.data2)
```

**Q4: Compare old and new data**

We will use the same calculation strategy, except for the new set.

```{r}
mean.steps2.per.day <- mean(step.data2$steps, na.rm = TRUE)
median.steps2.per.day <- median(step.data2$steps, na.rm = TRUE)

step.data2 <- step.data2  %>% 
    arrange(date) %>%
    mutate(month.name=months(date), weekday=weekdays(date))
print(step.data2)    

totalsteps2.per.day <- step.data2  %>% 
    group_by(date) %>%
    summarize(steps.total=sum(steps))
# print (totalsteps.per.day)
sum.steps2.total <- sum(totalsteps2.per.day$steps.total)
mean.steps2.total <- mean(totalsteps2.per.day$steps.total)
median.steps2.total <- median(totalsteps2.per.day$steps.total)
```

> Make a histogram of the total number of steps taken each day 

``` {r}
ggplot(step.data2, aes(x = date, y = steps, fill = weekday)) + 
    geom_histogram(stat = 'identity') +
    ## add lines for the mean and the median - notice that being so close together they overlap
    geom_hline(aes(yintercept=mean.steps2.total), color = 'red', linetype='dashed') +
    geom_hline(aes(yintercept=median.steps2.total), color = 'black', linetype='dotted') +
    labs(title = 'Total number of steps taken each day', x = 'Date', y = 'Steps')
```

> Calculate and report  the **mean** and **median** total number of steps taken per day.

After applyig the fill-in strategy, we have an **average** of `r fd(mean.steps2.total)` steps 
and a **median** value of `r fd(median.steps2.total)` steps.

> Do these values differ from the estimates from the first part of the assignment?
What is the impact of imputing missing data on the estimates of the total daily number of steps?

|          |                         Mean |                           Median |
|---------:|-----------------------------:|---------------------------------:|
|Original  |      `r fd(mean.steps.total)`|        `r fd(median.steps.total)`|
|*New*     |   *`r fd(mean.steps2.total)`*|     *`r fd(median.steps2.total)`*|



## Are there differences in activity patterns between weekdays and weekends?

> Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

